Automatic discovery of cell types and microcircuitry from neural connectomics

### 2015-02-13 docs/paper.tex
updates to figure plots, etc

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=183mm]{fig1b.ai}}
  \caption{Deriving circuitry and types from connectomics data a. As input we take the connectivity between cells (b), the distance between them (c), the depth of the cell bodies (d), and the depth profile of the synapses (e). f. Our algorithm discovers hidden cell types in this
    connectivity data by assuming all cells of a type share a
    distance-dependent connectivity profile, similar depth, and a
    similar synaptic density profile, with cells of other types.  This
    results in a clustering of the cells by those hidden
    types. f.) shows the cell connectivity matrix with cells of the
    same type grouped together. g.) shows the learned probability of
    connection between our different types at various distances -- in
    this case, the cells are likely to connect when they are
    close. h. shows the probability of connection between two cell types
that very rarely connect -- there's a background ``base'' connection rate
to account for errors in data, but the probability is very low. 
We can plot this these types (i) to show the relationship and spatial
connections, and how probable various types are to connect to one
another. j.) The connectivity between our discovered types changes as a
function of distance. }

\label{fig:overview}
\end{figure}

@@

\subsection{Ground Truth Validation}
To validate our model, we performed a series of simulations to test if
the model can accurately recover the true underlying network structure
and cell type identity.  We thus simulate data for which we know the
correct structure and compare the estimated structure based on the
algorithm (see methods) with the one we used for simulation. We find
that the model does a good job of recovering the correct number of
cell types, (fig~\ref{fig:synthetic}a), the the cell identities
(fig~\ref{fig:synthetic}b), and the spatial extent of each type
(fig~\ref{fig:synthetic}c).  For comparison, the infinite
stochastic block model (fig~\ref{fig:synthetic}a, b, c, black line) assumes that only cell type matters, and thus finds
small neighborhoods of connected nodes (instead of global connectivity
patterns). Our model converges relatively quickly (see supplemental) to an estimate of
the most probable values for the cell types, which is enabled by using
a combination of simulated annealing and parallelized Markov chain
Monte Carlo (see methods for details). Thus our model at least is promising to be applied to biological datasets.

\subsection{Model Mismatch}
To understand how our model would perform on real datasets that do not perfectly adhere to its generative model we next want to analyze its performance in such cases. To show where our model
performs poorly, we attempt connectivity inference on five sets of
synthetic data. Our type-and-distance assumptions for our connectivity model are
simplifying, and reflect a general consensus about the organization of
some (but not all) types of neural tissue. We assume that type is discrete (as opposed to the lpcm). We assume that each neuron only has one type (as opposed to the mixed membership model). We assume that connectivity is a function of distance and type (as opposed to the classic stochastic block model). Our goal is to show that our model does not find
``structure'' when there is none.

We generate ten synthetic datasets based on assumptions of other models and analyze our model’s performance on these datasets. More specifically we perform 10-way cross validation to predict link accuracy as well as ARI to the known (``true'') type, type spatial extent, and type number. We find that x. We find that y. We find that z. Our results thus strongly suggest even when the data is not generated according to our own generative model that our results are meaningful.

\fxwarning{Add results} 
\fxwarning{add spatial extent of types}

\fxwarn{add to figure comments}
\begin{itemize}
\item Data with type-specific connectivity and distance dependence
\item Data with type-specific connectivity and no distance dependence
\item Data generated according to a latent-position clustering model,
  where each cell belongs to a type which is part of a cluster in a
  low-dimensional space \fxwarning{Cite this paper}
\item cell type is combinatorial -- there are three possible
  properties a cell can have, and each subset of properties influences
  connectivity to neighboring cells. 
\item cell type is a continuum -- each cell's type continuously varies
  from zero to one, and the probability of a connection between cells
  depends on this type.
\end{itemize}

@@

@@

Having shown our model to work on the repeating tessellated, laminar structure of the mammilian retina, we then apply our model to a structurally very different connectome -- the whole body of a small roundworm: 
\textit{Caenorhabditis elegans} is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons developmentally determined,
leading to early measurement of the connectome. Unlike the retina,
only the motor neurons in C. elegans exhibit regular distribution in
space -- along the body axis. Most interneurons are
concentrated in various ganglia that project throughout the entire
animal, and the sensory neurons are primarily located in a small
number of anterior ganglia. C. elegans also differs from the retina in that the measured connectome is actually two separate graphs -- one of directed chemical synapses and another of undirected electrical synapses.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=4.5in]{celegans.ai}}
  \caption{Discovering connectivity and type in \textit{C. elegans}. 
 a.) Initial
    \textit{C. elegans} cell connectivity matrix, red are
    (directional) chemical synapses, blue are electrical
    synapses. Point size is proportional to synapse count between
    those cells. b.) Clustered adjacency matrix showing identified
    clusters. c.) Inter-cluster connectivity, showing the position of
    each cell along the body axis as well as the ``true'' cell type --
    both coarse (motor, sensory, and interneuron) as well as fine
    (small text labels). }
  \label{fig:celegans}
\end{figure}

Using both the chemical and electrical connectivity (see methods), we
determined the underlying clusters explained by connectivity and
distance (fig~\ref{fig:celegans}a). A superficial inspection of the results shows
clustering into groups consisting roughly homogeneously of motor
neurons, sensory neurons, and interneurons. Closer examination reveals
agreement with the classifications originally outlined by White in
1986.

We identify cell types that reflect the known motor/non-motor neuron
classes, even though this system lacks the strong repeat
microcircuitry our model was designed for.  Motor-neuron types AS, DA,
and VA, all exclusively postsynaptic, are identified as a common type,
as are motor-neuron types VD and DD. Traditional types VC, DB, and VB
also mostly share a cluster. Various head motor neurons, including
types SMD and RMD, are clustered together. Interneurons with known
anatomically-distinct connectivity patterns, such as AVA (2 cells),
are clustered into pure types.The algorithm even correctly places the
single-cell types DVB and DVC by themselves.

@@

\section*{Discussion}
We have presented a machine learning technique that allows cell types
and microcircuitry to be discovered from connectomics data.  We have
shown its applicability to regularly structured laminar neural
circuits like the retina, as well as a less structured whole neuronal
organism (\textit{C. elegans}) and an artificial processor. When
compared to existing methods, we show how the incorporation of all of
this data yields results that combine both high link-prediction
accuracy and high agreement with human anatomists. We have found that
combining the available data types allows us to discover cell types
and microcircuitry that were known to exist in the systems based on
decades of previous research and allows good prediction of
connectivity.

@@

For our probabilistic model, inference becomes slower as the amount of
data increases. Our algorithm required several hours for 1000
neurons. Scaling this class of probabilistic model is an active area
of research, and recent results in both variational methods
\autocite{Hoffman2013} and spectral learning \autocite{Anandkumar2012}
and future work could adapte them to find faster approximate solutions
to our problem.

>>>>

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=183mm]{fig1b.ai}}
  \caption{Deriving circuitry and types from connectomics data a. As
    input we take the connectivity between cells (b), the distance
    between them (c), the depth of the cell bodies (d), and the depth
    profile of the synapses (e). f. Our algorithm discovers hidden
    cell types in this connectivity data by assuming all cells of a
    type share a distance-dependent connectivity profile, similar
    depth, and a similar synaptic density profile, with cells of other
    types.  This results in a clustering of the cells by those hidden
    types. f.) shows the cell connectivity matrix with cells of the
    same type grouped together. g.) shows the learned probability of
    connection between our different types at various distances -- in
    this case, the cells are likely to connect when they are
    close. h. shows the probability of connection between two cell
    types that very rarely connect -- there's a background ``base''
    connection rate to account for errors in data, but the probability
    is very low.  We can plot this these types (i) to show the
    relationship and spatial connections, and how probable various
    types are to connect to one another. j.) The connectivity between
    our discovered types changes as a function of distance. }
\label{fig:overview}
\end{figure}

@@

\subsection{Validation with simulated data where ground truth is known}
To validate our model, we performed a series of simulations to test if
the model can accurately recover the true underlying network structure
and cell type identity.  We thus simulate data for which we know the
correct structure and compare the estimated structure based on the
algorithm (see methods) with the one we used for simulation. We find
that the model does a good job of recovering the correct number of
cell types, (fig~\ref{fig:synthetic}a), the the cell identities
(fig~\ref{fig:synthetic}b), and the spatial extent of each type
(fig~\ref{fig:synthetic}c).  For comparison, we show the results from instead using the infinite
stochastic block model (fig~\ref{fig:synthetic}a, b, c, black line) which assumes that only cell type matters, and thus finds
small neighborhoods of connected nodes (instead of global connectivity
patterns). This contrast shows that while the regular block model can not correctly deal with distance dependent connectivity, our model can. Our model converges relatively quickly (see supplemental) to an estimate of
the most probable values for the cell types, which is enabled by using
a combination of simulated annealing and parallelized Markov chain
Monte Carlo (see methods for details). Thus our model at least is promising to be applied to biological datasets.

\subsection{Model Mismatch}

\fxwarning{Add xlabel, ylabel, make line width of axes a bit bigger, larger number font, 
all label fonts should be the same size }

To understand how our model would perform on real datasets that do not
perfectly adhere to its generative model we next want to analyze its
performance in such cases. To understand the properties of our model, we
attempt connectivity inference on four sets of synthetic data. This helps us understand what our model would do if the data does not obey our assumptions.

We thus generate ten sets of synthetic data from each of four existing models. 
The distance-dependent stochastic block model assumes type depends on distance, 
the traditional stochastic block model has no notion of distance, the mixed membership block model assumes type is combinatorial, and the latent position cluster model assumes that type is clustered-but-continuous.

If the data is sampled from our model, inference according to our model, unsurprisingly, is good by all measures. It correctly estimates the number of cell types; it is good at predicting connectivity (high AUC); it agrees with human classification (Rand index); it discovers all types, and leads to homogenous estimates (Fig. \ref{fig:othermodel} first row). If the data comes from a block model without distance dependence, we see that it still does pretty well on all meaningful measures (Fig. \ref{fig:othermodel} second row). It basically estimates correctly that there is no distance dependence. For the mixed membership model (Fig. \ref{fig:othermodel} third row), the model grossly overestimates the number of types, by basically allocating a type for each combination of memberships. Otherwise it still performs relatively well. Lastly, for the latent position clustering model (Fig. \ref{fig:othermodel} fourth row), the model does poorly. If type is continuous instead of discrete then our model is basically trying to cover a continuous set with a discrete scenario leading to rather poor performance. However, as we do expect cell types to be real in biology we might expect our model to do well with real data.

\begin{figure}
  \centering

  \includegraphics[width=6.0in]{synthdifferent.ai}
  \caption{Model inferences when true generating model differs from our distance-block-model prior. Horizontal columns show results with synthetic data generated according to the latent position cluster model, the mixed membership block model, the non-distance dependent stochastic block model, and the distance dependent stochastic block model. In all cases histograms represent posterior distribution over indicated metric. a.) The number of types found by the model, vertical dash line indicates ``true'' type number (not applicable to mixed membership model). b.) The area under the ROC curve, indicating link prediction accuracy. c, d, e.) Clustering metrics
quantifying degree of type agreement with known ground truth. }

  \label{fig:othermodel}
\end{figure}

@@

\begin{figure}
  \centering 
  % generated from experiments/edgeeffects/edgeeffects.ipynb
  \centerline{\includegraphics[width=3.5in]{edgeeffects.pdf}}
  \caption{edge effects}
\label{fig:synthetic}
\end{figure}

\fxwarning{cut down on the number of ticks, this will be a small figure, homogenize fonts}

@@

Having shown our model to work on the repeating tessellated, laminar structure of the mammalian retina, we then apply our model to a structurally very different connectome -- the whole body of a small roundworm: 
\textit{Caenorhabditis elegans} is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons developmentally determined,
leading to early measurement of the connectome. Unlike the retina,
only the motor neurons in C. elegans exhibit regular distribution in
space -- along the body axis. Most interneurons are
concentrated in various ganglia that project throughout the entire
animal, and the sensory neurons are primarily located in a small
number of anterior ganglia. \textit{C. elegans} also differs from the retina in that the measured connectome is actually two separate graphs -- one of directed chemical synapses and another of undirected electrical synapses. As this is a very different connectome, it allows an interesting generalization test: how well will our model work on such a distinct data set.

\fxwarning{Fix linewidth too small, align d.e.f. callouts, change multiple fonts. Boxes same size. Need y labels (number of samples in bin) }

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=4.7in]{celegans2.ai}}
  \caption{Discovering connectivity and type in
    \textit{C. elegans}. Discovered types a.) The cell/cell
    coassignment matrix identifies the posterior probability of any
    two cells in \textit{c. elegans} being of the same type. The large
    visible clusters roughly correspond to a more-coarse version of
    types identified by neuroanatomists seen in b.) In particular,
    robust subpopulations are found corresponding to divisions of
    moter, interneuron, and sensory neurons. c.) The location of the
    cell soma, revealing how spatial localizatinion correlates with
    type. d.) The ROC curves for held-out link probability for both
    the electrical synapses (gap junctions) and chemical synapses in
    c. elegans. e.) The posterior distribution of the area under the
    ROC curve for the curves in d. f.) Measurements of the
    agreement of our identified cell types compared to neuroanatomists. 
    See methods for an explanation of metrics. }
  \label{fig:celegans}
\end{figure}

Using both the chemical and electrical connectivity (see methods), we
determined the underlying cell types explained by connectivity and
distance (fig~\ref{fig:celegans}a). A superficial inspection of the results shows
clustering into groups consisting roughly homogeneously of motor
neurons, sensory neurons, and interneurons. Closer examination reveals
agreement with the classifications originally outlined by White in
1986.

@@

\fxwarning{Add chip back in}

\section*{Discussion}
We have presented a machine learning technique that allows cell types
and microcircuitry to be discovered from connectomics data.  We have
shown its applicability to regularly structured laminar neural
circuits like the retina, as well as a less structured whole neuronal
organism (\textit{C. elegans}) and a classical processor. When
compared to existing methods, we show how the incorporation of all of
this data yields results that combine both high link-prediction
accuracy and high agreement with human anatomists. We have found that
combining the available data types allows us to discover cell types
and microcircuitry that were known to exist in the systems based on
decades of previous research and allows good prediction of
connectivity.

@@

For our probabilistic model, inference becomes slower as the amount of
data increases. Our algorithm required several hours for 1000
neurons. Scaling this class of probabilistic model is an active area
of research, and recent results in both variational methods
\autocite{Hoffman2013} and spectral learning \autocite{Anandkumar2012}
and future work could adapt them to find faster approximate solutions
to our problem.
